Adaptive Exponentiated Gradient

### 2014-01-26 lp/ftrl-aux.tex
done with matrix section
fixed missing references

\section{Choosing $m_t$ via Follow the Regularized Leader} 
\label{sec:ftrl-aux}
In the previous section we introduced an adaptive exponentiated 
gradient algorithm that achieved a regret bound of 
\[ \frac{\log(n)}{\eta} + \eta \sum_{i=1}^n w_i \sum_{t=1}^T (z_{t,i} - m_{t,i})^2 \]
for any sequence $m_t$. We now turn our attention to the task of 
optimizing $m_t$. Clearly, the optimal value of $m_t$ is $m_t = z_t$. However, 
this would require perfect knowledge of the future. Suppose we are less ambitious 
and only want to be competitive with the best fixed value of $m_t$. This optimal value 
is $m^* \eqdef \frac{1}{T} \sum_{t=1}^T z_t$, but we still have the issue that 
we do not $m^*$ ahead of time. However, we have now cast the problem in the familiar 
regret minimzation framework, and can use the machinery of follow the regularized 
leader:
\begin{proposition}
\label{prop:ftrl-aux}
Suppose that we choose $m_{t,i} = \frac{1}{t} \sum_{s=1}^{t-1} z_{s,i}$. 
Then, for all $i$, 
\[ \sum_{t`=1}^T (z_{t,i} - m_{t,i})^2 \leq \sum_{t=1}^T (z_{t,i} - m^*_i)^2 + 4\log(T+1). \]
\end{proposition}
\begin{proof}
TODO
\end{proof}
Combining Proposition~\ref{prop:ftrl-aux} with Proposition~\ref{prop:aeg} yields 
the algorithm titled EG-Variance in Table~\ref{tab:algorithms}. This algorithm has the 
same updates as an algorithm given by \cite{hazan2010variation}, but uses the 
optimistic prediction $\exp(\theta_t - \eta m_t)$ rather than $\exp(\theta_t)$. 
As a straightforward consequence of Propositions~\ref{prop:aeg} and~\ref{prop:ftrl-aux}, we 
obtain the following regret bound, which improves the bound presented in 
\cite{hazan2010variation}:
\begin{corollary}
The EG-Variance algorithm achieves a regret bound of
%\[ \Regret(w) \leq \frac{1 + (\log(n)-1)\|w\|_1 + \sum_{i=1}^n w_i\log(w_i)}{\eta} + \eta \left[ \sum_{i=1}^n w_i \sum_{t=1}^T (z_{t,i} - m_i^*)^2 + 4\|w\|_1\log(T) \right]. \]
\[ \Regret(w) \leq \frac{\log(n)}{\eta} + \eta \left[ \sum_{i=1}^n w_i \sum_{t=1}^T (z_{t,i} - m_i^*)^2 + 4\log(T) \right]. \]
\end{corollary}

>>>>

\section{Choosing $m_t$ via Follow the Regularized Leader} 
\label{sec:ftrl-aux}
In the previous section we introduced an adaptive exponentiated 
gradient algorithm that achieved a regret bound of 
\[ \frac{\log(n)}{\eta} + \eta \sum_{i=1}^n w_i \sum_{t=1}^T (z_{t,i} - m_{t,i})^2 \]
for any sequence $m_t$. We now turn our attention to the task of 
optimizing $m_t$. Clearly, the optimal value of $m_t$ is $m_t = z_t$. However, 
this would require perfect knowledge of the future. Suppose we are less ambitious 
and only want to be competitive with the best fixed value of $m_t$. This optimal value 
is $m^* \eqdef \frac{1}{T} \sum_{t=1}^T z_t$, but we still have the issue that 
we do not $m^*$ ahead of time. However, we have now cast the problem in the familiar 
regret minimzation framework, and can use the machinery of follow the regularized 
leader \cite{hazan2011ftrl}:
\begin{proposition}
\label{prop:ftrl-aux}
Suppose that we choose $m_{t,i} = \frac{1}{t} \sum_{s=1}^{t-1} z_{s,i}$. 
Then, for all $i$, 
\[ \sum_{t=1}^T (z_{t,i} - m_{t,i})^2 \leq \sum_{t=1}^T (z_{t,i} - m^*_i)^2 + 4\log(T+1). \]
\end{proposition}
\begin{proof}
TODO
\end{proof}
Combining Proposition~\ref{prop:ftrl-aux} with Proposition~\ref{prop:aeg} yields 
the algorithm titled EG-Variance in Table~\ref{tab:algorithms}. This algorithm has the 
same updates as an algorithm given by \cite{hazan2010variation}, but uses the 
optimistic prediction $\exp(\theta_t - \eta m_t)$ rather than $\exp(\theta_t)$. 
As a straightforward consequence of Propositions~\ref{prop:aeg} and~\ref{prop:ftrl-aux}, we 
obtain the following regret bound, which improves the bound presented in 
\cite{hazan2010variation}:
\begin{corollary}
\label{cor:EG-variance}
The EG-Variance algorithm achieves a regret bound of
%\[ \Regret(w) \leq \frac{1 + (\log(n)-1)\|w\|_1 + \sum_{i=1}^n w_i\log(w_i)}{\eta} + \eta \left[ \sum_{i=1}^n w_i \sum_{t=1}^T (z_{t,i} - m_i^*)^2 + 4\|w\|_1\log(T+1) \right]. \]
\[ \Regret(w) \leq \frac{\log(n)}{\eta} + \eta \left[ \sum_{i=1}^n w_i \sum_{t=1}^T (z_{t,i} - m_i^*)^2 + 4\log(T+1) \right]. \]
\end{corollary}

### 2014-01-26 lp/introduction.tex
fixed missing references

\section{Introduction} 
\label{sec:intro}
The exponentiated gradient algorithm is a powerful tool for performing online 
learning in the presence of many irrelevant features 
\cite{kivinen1997, littlestone1988}. One common application of exponentiated 
gradient is to the learning from experts setting, in which it is also known 
as the weighted majority algorithm \cite{wm} and entertains regret bounds of the 
form 
\[ \frac{\log(n)}{\eta} + \eta \sum_{i=1}^n \sum_{t=1}^T w_{t,i}z_{t,i}^2, \]
where $w_t$ is the weight vector of the learner at time $t$ and $z_t$ is the 
vector of losses. Such bounds can be obtained under the mirror descent 
framework using a local norms analysis \cite{shalev2011}.

@@

We achieve such a regret bound by combining two ideas: the first is 
Rakhlin and Sridharan's framework of \emph{optimistic learning} 
\cite{rakhlin2012}, which in our setting yields bounds of the form 
\begin{equation}
\label{eqn:optimistic-regret}
\frac{\log(n)}{\eta} + \eta \sum_{i=1}^n w_i \sum_{t=1}^T (z_{t,i} - m_{t,i})^2.
\end{equation}
If we could set $m_t$ to be $m^*$, then we would achieve the bound in 
(\ref{eqn:variance-regret}). Unfortunately, $m^*$ is only known after observing 
the full sequence of gradients. To overcome this, we learn $m^*$ itself in an 
online fashion, choosing the $m_t$ according to a Follow the Regularized Leader 
(FTRL) algorithm \cite{ftrl}. The regret bound thus obtained is given in 
Theorem~\ref{thm:variance-vector}, which improves a result by 
\cite{hazan2010variation}.

Finally, we extend all these results to the matrix setting, where $W_t$ is now 
a positive semidefinite matrix, possibly constrained to have trace $1$ 
(in analogy with the simplex). This setting has already been extensively studied 
\cite{lotsofpeople} and is important in obtaining online bounds for various 
combinatorial optimization problems \cite{morepeople}. Using the machinery so 
far developed, all of our results extend straightforwardly to the matrix setting, 
with the exception of the auxiliary FTRL algorithm for estimating $M^*$. In this 
case we need a new algorithm and analysis tool: a variant of FTRL for 
\emph{vector-valued} losses ordered relative to some cone $\sK$. This generalized 
FTRL lemma is given in Lemma~\ref{lem:ftrl-k} and may be of independent interest.

@@

Other notions of ``adaptive'' weights have also been explored; though unrelated 
to the current work, we briefly mention them for completeness. The first notion 
notion of adaptivity pertains to time-varying experts: here the desire is to 
obtain regret relative to the best \emph{schedule} of experts, where the number 
of switches in the schedule is bounded. This is explored in \cite{switching}. 
The other slightly more related notion is an adaptive step size, where the step 
size $\eta$ changes as a function of time (often at a fixed schedule of 
$1/\sqrt{T}$, but sometimes in a more sophisticated fashion). This idea is used 
extensively throughout the literature but a few particularly relevant papers are 
\cite{adaptive-step}.

Finally, we note that the exponentiated gradient itself has been analyzed 
and applied in various settings. One common setting, using a hinge-like loss, 
is the Winnow Algorithm \cite{winnow}, which has been carefully analyzed in 
\cite{winnow-careful}.

>>>>

\section{Introduction} 
\label{sec:intro}
The exponentiated gradient algorithm is a powerful tool for performing online 
learning in the presence of many irrelevant features 
\cite{kivinen1997, littlestone1988}. One common application of exponentiated 
gradient is to the learning from experts setting, in which it is also known 
as the weighted majority algorithm \cite{littlestone1989wm} and entertains regret bounds of the 
form 
\[ \frac{\log(n)}{\eta} + \eta \sum_{i=1}^n \sum_{t=1}^T w_{t,i}z_{t,i}^2, \]
where $w_t$ is the weight vector of the learner at time $t$ and $z_t$ is the 
vector of losses. Such bounds can be obtained under the mirror descent 
framework using a local norms analysis \cite{shalev2011}.

@@

We achieve such a regret bound by combining two ideas: the first is 
Rakhlin and Sridharan's framework of \emph{optimistic learning} 
\cite{rakhlin2012}, which in our setting yields bounds of the form 
\begin{equation}
\label{eqn:optimistic-regret}
\frac{\log(n)}{\eta} + \eta \sum_{i=1}^n w_i \sum_{t=1}^T (z_{t,i} - m_{t,i})^2.
\end{equation}
If we could set $m_t$ to be $m^*$, then we would achieve the bound in 
(\ref{eqn:variance-regret}). Unfortunately, $m^*$ is only known after observing 
the full sequence of gradients. To overcome this, we learn $m^*$ itself in an 
online fashion, choosing the $m_t$ according to a Follow the Regularized Leader 
(FTRL) algorithm \cite{hazan2011ftrl}. The regret bound thus obtained is given in 
Corollary~\ref{cor:EG-variance}, which improves a result by 
\cite{hazan2010variation}.

Finally, we extend all these results to the matrix setting, where $W_t$ is now 
a positive semidefinite matrix, possibly constrained to have trace $1$ 
(in analogy with the simplex). This setting has already been extensively studied 
\cite{tsuda2005matrix, arora2007combinatorial} and is important in 
obtaining online and approximation bounds for various combinatorial optimization 
problems \cite{arora2007combinatorial, hazan2012near}. Using the machinery so 
far developed, all of our results extend straightforwardly to the matrix setting, 
with the exception of the auxiliary FTRL algorithm for estimating $M^*$. In this 
case we need a new algorithm and analysis tool: a variant of FTRL for 
\emph{vector-valued} losses ordered relative to some cone $\sK$. This generalized 
FTRL lemma is given in Lemma~\ref{lem:ftrl-k} and may be of independent interest.

@@

Other notions of ``adaptive'' weights have also been explored; though unrelated 
to the current work, we briefly mention them for completeness. The first 
notion, which has occasionally been referred to as adaptivity, pertains to time-varying experts: here the desire is to obtain regret relative to the best \emph{schedule} of experts, 
where the schedule is in some way restricted to drift slowly over time 
\cite{auer1998tracking, kleinberg2010regret, chiang2012online}. 
The other slightly more related notion is an adaptive step size, where the step 
size $\eta$ changes as a function of time (often at a fixed schedule of 
$1/\sqrt{T}$, but sometimes in a more sophisticated fashion). This idea is used 
extensively throughout the literature but a few particularly relevant papers are 
\cite{hazan2007adaptive, orabona2013dimension}.

Finally, we note that the exponentiated gradient algorithm itself has been analyzed 
and applied in various settings. One common setting, using a hinge-like loss, 
is the Winnow Algorithm \cite{littlestone1988}, which has been carefully analyzed in 
\cite{sabato2012learning}.

### 2014-01-26 lp/mw12.tex
fixed missing references

\section{Two Types of Updates} 
\label{sec:mw12}
Our point of departure is the two different types of multiplicative updates 
mentioned in the introduction. For simplicity we will consider the setting of 
learning from experts, although our later theorems will hold in greater 
generality. In this setting there are $n$ experts, and the learner maintains a 
probability distribution $w_t \in \Delta_n$ over the experts, which gets 
updated in each of $T$ rounds. After selecting the distribution $w_t$, 
the loss vector $w_t \in [-1,1]^n$ of each of the experts is revealed, and the 
learner incurs loss $w_t^{\top}z_t$. The learner's goal is to minimize 
the regret $\sup_{w \in \Delta_n} \Regret(w)$, where
\[ \Regret(w) \eqdef \sum_{t=1}^T w_t^{\top}z_t - \sum_{t=1}^T w^{\top}z_t. \]
We consider two types of updates for the weight vector $w_t$. In both cases 
$w_1$ is initialized as $w_{1,i} = \frac{1}{n}$ for $1 \leq i \leq n$, but 
the updates differ, as in (\ref{eqn:mw1}) and (\ref{eqn:mw2}) below:
\begin{align}
\label{eqn:mw1}
w_{t+1,i} &\propto w_{t,i}\exp(-\eta z_{t,i})\\
\label{eqn:mw2}
w_{t+1,i} &\propto w_{t,i}(1-\eta z_{t,i})
\end{align}
The regret bounds for each of (\ref{eqn:mw1}) and (\ref{eqn:mw2}) are 
well-known \cite{people}, but we include them below for completeness.
\begin{theorem}
\label{thm:mw12}
For any $0 < \eta \leq \frac{1}{2}$, the updates (\ref{eqn:mw1}) and 
(\ref{eqn:mw2}) enjoy regret, respectively,
\begin{align}
\label{eqn:mw1-regret}
\Regret(w) &\leq \frac{\log(n)}{\eta} + \eta \sum_{i=1}^n \sum_{t=1}^T w_{t,i}z_{t,i}^2 \\
\label{eqn:mw2-regret}
\Regret(w) &\leq \frac{\log(n)}{\eta} + \eta \sum_{i=1}^n w_i \sum_{t=1}^T z_{t,i}^2
\end{align}
\end{theorem}
To understand why (\ref{eqn:mw2-regret}) may be a better bound than 
(\ref{eqn:mw1-regret}), suppose that the best expert has a loss identically 
equal to zero.\footnote{Of course, if we knew that this was the case ahead of 
time, there would be far better algorithms; we use this scenario purely for 
illustrative purposes.} Then the optimal $w$ places all mass on this expert, 
and (\ref{eqn:mw2-regret}) reduces to $\frac{\log(n)}{\eta} = 2\log(n)$ for 
$\eta = \frac{1}{2}$.

>>>>

\section{Two Types of Updates} 
\label{sec:mw12}
Our point of departure is the two different types of multiplicative updates 
mentioned in the introduction. For simplicity we will consider the setting of 
learning from experts, although our later theorems will hold in greater 
generality. In this setting there are $n$ experts, and the learner maintains a 
probability distribution $w_t \in \Delta_n$ over the experts, which gets 
updated in each of $T$ rounds. After selecting the distribution $w_t$, 
the loss vector $w_t \in [-1,1]^n$ of each of the experts is revealed, and the 
learner incurs loss $w_t^{\top}z_t$. The learner's goal is to minimize 
the regret $\sup_{w \in \Delta_n} \Regret(w)$, where
\[ \Regret(w) \eqdef \sum_{t=1}^T w_t^{\top}z_t - \sum_{t=1}^T w^{\top}z_t. \]
We consider two types of updates for the weight vector $w_t$. In both cases 
$w_1$ is initialized as $w_{1,i} = \frac{1}{n}$ for $1 \leq i \leq n$, but 
the updates differ, as in (\ref{eqn:mw1}) and (\ref{eqn:mw2}) below:
\begin{align}
\label{eqn:mw1}
w_{t+1,i} &\propto w_{t,i}\exp(-\eta z_{t,i})\\
\label{eqn:mw2}
w_{t+1,i} &\propto w_{t,i}(1-\eta z_{t,i})
\end{align}
The regret bounds for each of (\ref{eqn:mw1}) and (\ref{eqn:mw2}) are 
well-known \cite{shalev2011, cesa2007}, but we include them below for completeness.
\begin{theorem}
\label{thm:mw12}
For any $0 < \eta \leq \frac{1}{2}$, the updates (\ref{eqn:mw1}) and 
(\ref{eqn:mw2}) enjoy regret, respectively,
\begin{align}
\label{eqn:mw1-regret}
\Regret(w) &\leq \frac{\log(n)}{\eta} + \eta \sum_{i=1}^n \sum_{t=1}^T w_{t,i}z_{t,i}^2 \\
\label{eqn:mw2-regret}
\Regret(w) &\leq \frac{\log(n)}{\eta} + \eta \sum_{i=1}^n w_i \sum_{t=1}^T z_{t,i}^2
\end{align}
\end{theorem}
To understand why (\ref{eqn:mw2-regret}) may be a better bound than 
(\ref{eqn:mw1-regret}), suppose that the best expert has a loss identically 
equal to zero.\footnote{Of course, if we knew that this was the case ahead of 
time, there would be far better algorithms; we use this scenario purely for 
illustrative purposes.} Then the optimal $w$ places all mass on this expert, 
and (\ref{eqn:mw2-regret}) reduces to $\frac{\log(n)}{\eta} = 2\log(n)$ for 
$\eta = \frac{1}{2}$.
